---
title: "Workflow: basics"
author: "Haris Farooqi"
format: 
  html:
    embed-resources: true
editor: visual
---

**2.5 Exercises**

1.  Why does this code not work?

```{r}
my_variable <- 10
my_varıable
#> Error in eval(expr, envir, enclos): object 'my_varıable' not found
```

-   my_variaable was spelt incorrectly.

2.  Tweak each of the following R commands so that they run correctly:

```{r}
#incorrect code:
#libary(todyverse)

#ggplot(dTA = mpg) + 
#  geom_point(maping = aes(x = displ y = hwy)) +
#  geom_smooth(method = "lm)


library(tidyverse)

ggplot(mpg, aes(x=displ,y=hwy)) + 
  geom_point() +
  geom_smooth(method="lm")
```

3.  Use Alt + Shift + K to open keyboard menu in RStudio.
4.  Let’s revisit an exercise from the [Section 1.6](https://r4ds.hadley.nz/data-visualize#sec-ggsave). Run the following lines of code. Which of the two plots is saved as `mpg-plot.png`? Why?

```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

-   my_bar_plot is saved since we have stored the plot we created in the my_bar_plot variable. then, we reference it in the argument plot in ggsave().
